\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2025-06-25 doc/jss-accepted-version/tex/jss4956.tex
Switch to pak::pak
Update tex output with updated packages/text

In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts by grouping factors (Figure~\ref{fig:diagram}).
All these components are additive in link space.
Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a more complex model could be written as
\[
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma}_t +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\alpha_g +
O_{\bm{s},t} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right).
\end{aligned}
\]
where the $\bm{X}$ represent model matrices, $\bm{\gamma}_t$ represents a vector of coefficients that are constrained to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O_{\bm{s},t}$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
We demonstrate these model components in Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}.

@@

The \sdmTMB\ formula syntax (\code{formula} argument) follows a standard format that is similar to \pkg{glmmTMB}, \pkg{lme4}, and \pkg{mgcv}.
In addition to standard main effects, the user can include random intercepts (e.g., \code{+ (1 | group)}), threshold shaped hockey-stick models \fct{breakpt} \citep{barrowman2000}, logistic functions \fct{logistic}, and penalized smoothers \fct{s} for generalized additive models, GAMS \citep{wood2017a}.
Penalized smoothers use the same \fct{s} and \fct{t2} syntax as in \pkg{mgcv} \citep{wood2017a}.
Supported functionality includes bivarate smoothers \code{s(x, y)},
smoothers varying by continuous or categorical variables \code{s(x1, by = x2)},
cyclical smoothers \code{s(x, bs = "cc")},
and smoothers with specified basis dimensions \code{s(x, k = 4)} \citep{wood2017a}.
Beyond the main formula, \fct{sdmTMB} accepts one-sided formulas for coefficients that should vary through time (\code{time\_varying}) according to a random walk or AR(1) process (\code{time\_varying\_type}) or vary through space as random fields (\code{spatial\_varying}).

Once the user makes a call to \fct{sdmTMB}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} \citep{Matrix} as an interface to the \pkg{Eigen} library in \proglang{C++}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using A6 \fct{stats::optimHess} \citep{gay1990, r2024}.
Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
it is also possible to apply an ``epsilon'' estimator \citep{thorson2016bias}, which corrects for bias arising from the variance and skewness of random effects when calculating an estimator as a non-linear transformation or random effects (see Section \ref{index-bias-correct} for an example).
Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristensen2016, zheng2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to sample from the joint posterior distribution for Bayesian inference.

@@

Prediction on fitted or new data is accomplished with \fct{predict} (\code{?predict.sdmTMB}).
In this paragraph, we include relevant \fct{predict} arguments in parentheses.
The predict method is flexible and includes the option to specify a new data frame (\code{newdata}),
whether to return predictions on the link or response scale (\code{type}),
whether to return standard errors (\code{se\_fit},
which can be slow if conditioned on random fields),
whether to condition on the random fields (\code{re\_form}),
whether to condition on the random intercepts (\code{re\_form\_iid}),
which delta model linear predictor to use (\code{model}),
whether to take draws from the joint parameter precision matrix (\code{nsim}),
and whether to use MCMC samples from a \pkg{tmbstan} model fit (\code{mcmc\_samples}).

@@

\begin{Schunk}
\begin{Sinput}
R> install.packages("pak")
R> pak::pkg_install("pbs-assess/sdmTMB", dependencies = TRUE)
\end{Sinput}
\end{Schunk}

@@

                     coef.est coef.se
(Intercept)             -0.44    0.42
poly(log(depth), 2)1   -75.72    9.30
poly(log(depth), 2)2  -109.79    9.27

@@

Delta/hurdle model 1: -----------------------------------
Family: binomial(link = 'log') 
            coef.est coef.se
(Intercept)     2.35    0.53
slog_depth      0.20    0.67

@@

Time-varying parameters:
                 coef.est coef.se
(Intercept)-2004     0.34    0.61
(Intercept)-2005     0.39    0.65
(Intercept)-2006     0.45    0.68
(Intercept)-2007     0.35    0.61
(Intercept)-2008     0.26    0.55
(Intercept)-2009     0.25    0.54
(Intercept)-2010     0.25    0.52
(Intercept)-2011     0.08    0.47
(Intercept)-2012    -0.09    0.49
(Intercept)-2013    -0.10    0.47
(Intercept)-2014    -0.10    0.45
(Intercept)-2015    -0.05    0.43
(Intercept)-2016     0.01    0.43
(Intercept)-2017    -0.01    0.45
(Intercept)-2018    -0.04    0.44
(Intercept)-2019    -0.15    0.47
(Intercept)-2020    -0.26    0.53
(Intercept)-2021    -0.38    0.60
(Intercept)-2022    -0.30    0.53

@@

Delta/hurdle model 2: -----------------------------------
Family: lognormal(link = 'log') 
            coef.est coef.se
(Intercept)     2.20    0.20
slog_depth      0.11    0.24

@@

Time-varying parameters:
                 coef.est coef.se
(Intercept)-2004     0.43    0.27
(Intercept)-2005     0.25    0.31
(Intercept)-2006     0.13    0.24
(Intercept)-2007     0.03    0.30
(Intercept)-2008    -0.06    0.26
(Intercept)-2009     0.20    0.30
(Intercept)-2010     0.50    0.26
(Intercept)-2011     0.30    0.30
(Intercept)-2012     0.18    0.23
(Intercept)-2013     0.10    0.29
(Intercept)-2014     0.04    0.23
(Intercept)-2015    -0.10    0.29
(Intercept)-2016    -0.26    0.23
(Intercept)-2017    -0.24    0.30
(Intercept)-2018    -0.27    0.23
(Intercept)-2019    -0.26    0.35
(Intercept)-2020    -0.30    0.35
(Intercept)-2021    -0.41    0.24
(Intercept)-2022    -0.20    0.25

@@

\begin{Schunk}
\begin{Sinput}
R> set.seed(42)
R> s <- simulate(fit, nsim = 500, type = "mle-mvn")
R> dharma_residuals(s, fit, test_uniformity = TRUE)
\end{Sinput}
\begin{figure}[ht]

@@

\caption[Simulation-based randomized quantile residuals from the \pkg{DHARMa} package]{Simulation-based randomized quantile residuals from the \pkg{DHARMa} package. The statistical test is a two-sided Kolmogorov-Smirnov test of the null hypothesis that the residuals are drawn from a uniform(0, 1) distribution.}\label{fig:dog-residuals}
\end{figure}
\end{Schunk}

@@

            coef.est coef.se
(Intercept)    -3.23    5.44
nao             0.36    0.22

Random intercepts:
       Std. Dev.
year_f       0.3

@@

Additional functionality in \sdmTMB\ not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model \citep{bakka2019}, and time-varying spatiotemporal covariance parameters \citep{ward2022}.
There are several planned future additions to the \sdmTMB\ model structure.
% Although we may introduce basic multivariate response modelling, we intend \sdmTMB\ to focus on univariate responses given \pkg{tinyVAST}'s focus on related models in a multivariate context.
A subset of features to be added include
(1) the ability to specify observation-specific likelihoods to integrate different data types \citep[e.g.,][]{gruss2019multiple},
(2) inclusion of a dispersion formula similar to \pkg{glmmTMB},
(3) inclusion of random slopes to complement the existing random intercepts,
and
(4) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base is written in \R\ rather than \proglang{C++}.
% (5) conditional autoregressive models as an alternative to random fields for large datasets that are already areal in nature.
% another feature that could be added is 3D models, e.g. https://doi.org/10.1016/j.spasta.2023.100750

>>>>

In practice, the above models can become considerably more complex by, for example, letting coefficients vary though time, letting coefficients vary through space \citep{hastie1993} as random fields, or adding random intercepts or slopes by grouping factors (Figure~\ref{fig:diagram}).
All these components are additive in link space.
Adopting the notation ``main'' for main effects, ``tvc'' for time-varying coefficients, and ``svc'' for spatially varying coefficients, a more complex model could be written as
\[
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma}_t +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\alpha_g +
O_{\bm{s},t} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right).
\end{aligned}
\]
where the $\bm{X}$ represent model matrices, $\bm{\gamma}_t$ represents a vector of coefficients that are constrained to vary through time as random walks or AR(1) processes, $\zeta_{\bm{s}}$ represents spatially varying coefficients following a random field, $\alpha_g$ represents IID random intercepts by group $g$ ($\alpha_g \sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right)$), $O_{\bm{s},t}$ represents an offset variable \citep[][p.~206]{mccullagh1989} (e.g., log sampling effort), and $\omega_{\bm{s}}$ and $\epsilon_{\bm{s},t}$ represent spatial and spatiotemporal intercept random fields as before (Figure~\ref{fig:diagram}).
We demonstrate these model components in Sections~\ref{pcod}, \ref{dogfish}, and \ref{svc-owls}.

@@

The \sdmTMB\ formula syntax (\code{formula} argument) follows a standard format that is similar to \pkg{glmmTMB}, \pkg{lme4}, and \pkg{mgcv}.
In addition to standard main effects, the user can include random intercepts and slopes (e.g., \code{+ (1 + x | group)}), threshold shaped hockey-stick models \fct{breakpt} \citep{barrowman2000}, logistic functions \fct{logistic}, and penalized smoothers \fct{s} for generalized additive models, GAMS \citep{wood2017a}.
Penalized smoothers use the same \fct{s} and \fct{t2} syntax as in \pkg{mgcv} \citep{wood2017a}.
Supported functionality includes bivarate smoothers \code{s(x, y)},
smoothers varying by continuous or categorical variables \code{s(x1, by = x2)},
cyclical smoothers \code{s(x, bs = "cc")},
and smoothers with specified basis dimensions \code{s(x, k = 4)} \citep{wood2017a}.
Beyond the main formula, \fct{sdmTMB} accepts one-sided formulas for coefficients that should vary through time (\code{time\_varying}) according to a random walk or AR(1) process (\code{time\_varying\_type}) or vary through space as random fields (\code{spatial\_varying}).

Once the user makes a call to \fct{sdmTMB}, input data structures for a \TMB\ model are constructed internally (Figure~\ref{fig:flowchart}).
If needed, data structures required to implement penalized smoothers are constructed with \fct{smooth2random} from \pkg{mgcv} \citep{wood2017a}, similarly to \pkg{gamm4} \citep{wood2020} and \pkg{brms} \citep{brms}.
\fct{sdmTMB} formats data, establishes parameter starting values, and constructs an objective function with derivatives based on a compiled \proglang{C++} template written for \TMB.
The objective function returns the marginal log likelihood and its gradient, integrating over random effects with the Laplace approximation \citep{kristensen2016} and efficient using sparse-matrix computation provided by \pkg{Matrix} \citep{Matrix} as an interface to the \pkg{Eigen} library in \proglang{C++}.
The negative marginal log likelihood is minimized via the non-linear optimization routine \fct{stats::nlminb} \citep{gay1990} in \R\ with additional optimization carried out via a Newton optimization step to find the Hessian using \fct{stats::optimHess} \citep{gay1990, r2024}.
Random effect values are returned at values that maximize the likelihood conditional on the fixed effects at their maximum marginal likelihood (i.e., plug-in or empirical Bayes estimates); however,
it is also possible to apply an ``epsilon'' estimator \citep{thorson2016bias}, which corrects for bias arising from the variance and skewness of random effects when calculating an estimator as a non-linear transformation or random effects (see Section \ref{index-bias-correct} for an example).
Standard errors on all parameters and derived quantities---including those involving random effects---are calculated using the generalized delta method \citep{kristensen2016, zheng2021}.
After rapid model exploration with maximum likelihood, one can optionally pass an \sdmTMB\ model to the \R\ package \pkg{rstan} \citep{carpenter2017, rstan} via \pkg{tmbstan} \citep{monnahan2018} to sample from the joint posterior distribution for Bayesian inference.

@@

Prediction on fitted or new data is accomplished with \fct{predict} (\code{?predict.sdmTMB}).
In this paragraph, we include relevant \fct{predict} arguments in parentheses.
The predict method is flexible and includes the option to specify a new data frame (\code{newdata}),
whether to return predictions on the link or response scale (\code{type}),
whether to return standard errors (\code{se\_fit},
which can be slow if conditioned on random fields),
whether to condition on the random fields (\code{re\_form}),
whether to condition on the random intercepts and slopes (\code{re\_form\_iid}),
which delta model linear predictor to use (\code{model}),
whether to take draws from the joint parameter precision matrix (\code{nsim}),
and whether to use MCMC samples from a \pkg{tmbstan} model fit (\code{mcmc\_samples}).

@@

\begin{Schunk}
\begin{Sinput}
R> install.packages("pak")
R> pak::pak("pbs-assess/sdmTMB", dependencies = TRUE)
\end{Sinput}
\end{Schunk}

@@

Conditional model:
                     coef.est coef.se
(Intercept)             -0.44    0.42
poly(log(depth), 2)1   -75.72    9.30
poly(log(depth), 2)2  -109.79    9.27

@@

Delta/hurdle model 1: -----------------------------------
Family: binomial(link = 'log')
Conditional model:
            coef.est coef.se
(Intercept)     2.35    0.53
slog_depth      0.20    0.67

@@

Time-varying parameters:
                 coef.est coef.se
(Intercept)-2004     0.34    0.61
(Intercept)-2005     0.39    0.65
(Intercept)-2006     0.45    0.68
(Intercept)-2007     0.35    0.61
(Intercept)-2008     0.26    0.55
(Intercept)-2009     0.25    0.54
(Intercept)-2010     0.25    0.52
(Intercept)-2011     0.08    0.47
(Intercept)-2012    -0.09    0.49
(Intercept)-2013    -0.10    0.47
(Intercept)-2014    -0.10    0.45
(Intercept)-2015    -0.05    0.43
(Intercept)-2016     0.01    0.43
(Intercept)-2017    -0.01    0.45
(Intercept)-2018    -0.04    0.44
(Intercept)-2019    -0.15    0.47
(Intercept)-2020    -0.26    0.53
(Intercept)-2021    -0.38    0.60
(Intercept)-2022    -0.30    0.53
rho-(Intercept)      0.85    0.28

@@

Delta/hurdle model 2: -----------------------------------
Family: lognormal(link = 'log')
Conditional model:
            coef.est coef.se
(Intercept)     2.20    0.20
slog_depth      0.11    0.24

@@

Time-varying parameters:
                 coef.est coef.se
(Intercept)-2004     0.43    0.27
(Intercept)-2005     0.25    0.31
(Intercept)-2006     0.13    0.24
(Intercept)-2007     0.03    0.30
(Intercept)-2008    -0.06    0.26
(Intercept)-2009     0.20    0.30
(Intercept)-2010     0.50    0.26
(Intercept)-2011     0.30    0.30
(Intercept)-2012     0.18    0.23
(Intercept)-2013     0.10    0.29
(Intercept)-2014     0.04    0.23
(Intercept)-2015    -0.10    0.29
(Intercept)-2016    -0.26    0.23
(Intercept)-2017    -0.24    0.30
(Intercept)-2018    -0.27    0.23
(Intercept)-2019    -0.26    0.35
(Intercept)-2020    -0.30    0.35
(Intercept)-2021    -0.41    0.24
(Intercept)-2022    -0.20    0.25
rho-(Intercept)      0.62    0.40

@@

\begin{Schunk}
\begin{Sinput}
R> set.seed(42)
R> s <- simulate(fit, nsim = 500, type = "mle-mvn")
R> dharma_residuals(s, fit)
\end{Sinput}
\begin{figure}[ht]

@@

\caption[Simulation-based randomized quantile residuals from the \pkg{DHARMa} package]{Simulation-based randomized quantile residuals from the \pkg{DHARMa} package.}\label{fig:dog-residuals}
\end{figure}
\end{Schunk}

@@

Random intercepts and/or slopes:

Conditional model:
     Groups        Name    Variance    Std.Dev.
     year_f (Intercept)        0.09        0.30

Conditional model:
            coef.est coef.se
(Intercept)    -3.23    5.44
nao             0.36    0.22

@@

Additional functionality in \sdmTMB\ not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model \citep{bakka2019}, and time-varying spatiotemporal covariance parameters \citep{ward2022}.
There are several planned future additions to the \sdmTMB\ model structure.
% Although we may introduce basic multivariate response modelling, we intend \sdmTMB\ to focus on univariate responses given \pkg{tinyVAST}'s focus on related models in a multivariate context.
A subset of features to be added include
(1) the ability to specify observation-specific likelihoods to integrate different data types \citep[e.g.,][]{gruss2019multiple},
(2) inclusion of a dispersion formula similar to \pkg{glmmTMB}, and
(3) integration with the \pkg{RTMB} \citep{rtmb} package so that the model code base is written in \R\ rather than \proglang{C++}.
% (5) conditional autoregressive models as an alternative to random fields for large datasets that are already areal in nature.
% another feature that could be added is 3D models, e.g. https://doi.org/10.1016/j.spasta.2023.100750
